---
title: How How to Change the Colors in a Sunburst Graph?
author: Thorn Thaler
date: '2022-03-04'
slug: how-how-to-change-the-colors-in-a-sunburst-graph
categories:
  - coding
tags:
  - plotly
  - visualization
authors: []
description: ''
externalLink: ''
series: []
editor_options: 
  chunk_output_type: console
---

```{r libs, include = FALSE}
library(glue)
```


## Sunburst Graphs

I was recently looking for a neat way to display nested ratios. What do I mean by 
*nested ratio*? Well, imagine a taxonomy where you have different hierarchical levels 
and you want to show the ratios of all levels simultaneously. 

There are a couple of different ways of how to show these, like mosaic plots, tree plots
or [sunburst charts](https://www.anychart.com/de/chartopedia/chart-type/sunburst-chart/).
Despite being a close relative to the most despised 
[pie chart](https://www.anychart.com/chartopedia/chart-type/pie-chart/) (of which Tufte
himself said that the *only worse design than a pie chart is several of them*^[*[The 
Visual Display of Quantitative Information, Edward Tufte, Graphics Press, 
2001](http://84.89.132.1/~michael/visualdata/tufte-aesthetics_and_technique.pdf)*]) they 
*can* be useful - if you use them with care-^[The same holds true for pie chart, if you use
them with 2 categories they *can* be informative, just because we are used to watches.
However, I side with Tufte here that a table or the bare number may be more informative in
these cases.]

Anyways, this should bot be blog post about the usefulness of pie charts and their 
relatives but the rtaher technical question of

1. How to ploit them?
1. How to change the color of the wedges?

So let's start with some code. I use the `Titanic` data set here, because it comes 
already with a nice hierarchical structure.

```{r get_data, skimr_include_summary = FALSE, message = FALSE}
library(dplyr)
library(forcats)
titanic <- as_tibble(Titanic) %>% 
   mutate(across(where(is.character), fct_inorder))
str(titanic)
```

You can see that we have 
`r glue("{nlevels(Class)} x {nlevels(Sex)} x {nlevels(Age)} x {nlevels(Survived)}", .envir = titanic)` rows corresponding to all possible combinations of 
`r glue_collapse(head(names(titanic), -1L), ", ", last = " and ")`.

The classical display (`?Titanic`) is the mosaic plot:

```{r mosaic-plot}
mosaicplot(Titanic)
```

Maybe it is me, but I find it difficult to read the graph properly. Furthermore, as the 
chart is static, we get only static information. Thus, if we want to get the underlying
numebrs we need to provide yet another table.

## Plotly

Enters `plotly`. With [plotly](https://plotly.com/r/) we can make our graphs more 
interactive and most notably we can create a sunburst chart. However, we need some
data aggregation first as the sunburt chart expects the data in a special format 
(c.f. [the docs](https://plotly.com/r/sunburst-charts/)). W.l.o.g. we define the
hierarchy as `Class > Age > Sex > Survived`. We have now to construct the hierarchy
using aggregation at different levels.

```{r sunburst}
library(plotly)
library(glue)

lvl4 <- titanic %>% 
   transmute(id = glue("{Class}-{Age}-{Sex}-{Survived}"),
             parent = glue("{Class}-{Age}-{Sex}"),
             value = n,
             label = Survived)

lvl3 <- titanic %>% 
   group_by(Class, Age, Sex) %>% 
   summarise(value = sum(n), .groups = "drop") %>% 
   transmute(id = glue("{Class}-{Age}-{Sex}"),
             parent = glue("{Class}-{Age}"),
             value,
             label = Sex)

lvl2 <- titanic %>% 
   group_by(Class, Age) %>% 
   summarise(value = sum(n), .groups = "drop") %>% 
   transmute(id = glue("{Class}-{Age}"),
             parent = glue("{Class}"),
             value,
             label = Age)

lvl1 <- titanic %>% 
   group_by(Class) %>% 
   summarise(value = sum(n), .groups = "drop") %>% 
   transmute(id = glue("{Class}"),
             parent = "Total",
             value,
             label = Class)

lvl0 <- titanic %>% 
   summarise(value = sum(n)) %>% 
   transmute(id = "Total",
             parent = "",
             value,
             label = "Total")

sunburst_data <- bind_rows(
   lvl0,
   lvl1,
   lvl2,
   lvl3,
   lvl4
)

(sb <- sunburst_data %>% 
   plot_ly() %>% 
   add_trace(ids = ~id,
             labels = ~ label,
             parents = ~ parent,
             values = ~ value,
             type = "sunburst",
             branchvalues = "total"))
```

Again, it may be personal preferences, but this chart seems to me much easier to 
understand:

1. We see that roughly 40% of all persons on the Titanic were crew members, 25% of the
passengers came 1st or 2nd class and a third from the 3rd class.
1. There were no children in the crew (ok, a no-brainer) and only a few women.
1. Females were more likely to survive than men.
1. Passengers from the better classes had higher survival rates.
1. Only children from the 3rd class died.

One nice property of the graph is that you can interactively zoom in and zoom out, which 
makes it very easy to read.

## Colors

Plotly has a nice default choice of colors, but you may want to change them. While it
seems to be straightforward for other plotly graphs it was not that clear to me how to 
change the color for the sunburst graph.

After a lot of googeling and more time on the 
[excellent reference guide for plotly](https://plotly.com/r/reference/) one path is to
overwrite the default color palette via 
[colorway](https://plotly.com/r/reference/#layout-colorway):


```{r sunburst-colorway}
library(viridis)
col_pal <- viridis(4)
sb %>% 
   layout(colorway = col_pal)
```

But what if we want to get even fainer control over individual segments say? Well, we
can specify a color for each segment individually, like this:

```{r sunburst-indiv-colors}
sunburst_data <- sunburst_data %>% 
   mutate(color = case_when(label == "Yes" ~ col_pal[1],
                            label == "No" ~ col_pal[3],
                            label == "Child" ~ "#C994C7",
                            label == "Adult" ~ "#DD1C77",
                            label == "Male" ~ "#377EB8",
                            label == "Female" ~ "#E41A1C",
                            label == "Crew" ~ "#FFFFCC",
                            label == "1st" ~ "#78C679",
                            label == "2nd" ~ "#C2E699",
                            label == "3rd" ~ "#238443",
                            TRUE ~ "#DFDFDF"))

sunburst_data %>% 
   plot_ly() %>% 
   add_trace(ids = ~id,
             labels = ~ label,
             parents = ~ parent,
             values = ~ value,
             marker = list(colors = ~ color),
             type = "sunburst",
             branchvalues = "total")

```

In this setting, the same categories get the same color. Admittedly, nobody with a 
clear mind would use such a colorful graph (not even those of us who grew up in the 80's 
like myself 😉), but you get my point.

Maybe a more realistic use case would be to highlight certain parts of the graph to guide
the reader's attention. For instance we could try to highlight the children on board of 
the titanic.

```{r sunburst-highlight-colors}
sunburst_data <- sunburst_data %>% 
   mutate(color = case_when(grepl("Child", id) ~ col_pal[1],
                            TRUE ~ "#DFDFDF"))

sunburst_data %>% 
   plot_ly() %>% 
   add_trace(ids = ~id,
             labels = ~ label,
             parents = ~ parent,
             values = ~ value,
             marker = list(colors = ~ color),
             type = "sunburst",
             branchvalues = "total")

```

*P.S.* One final remark: most probably you have observed that the sunburst graph uses some sort
of color intensity mapping, which results in reduced tints in the outer rings. I have
not yet found the setting which controls this behavior. If you know how to change this
setting, I will appreciate a quick comment.
