---
title: Aggregate Data with Nested Groups
author: Thorn Thaler
date: '2022-03-10'
slug: aggregate-data-with-nested-groups
categories:
  - coding
tags:
  - tidyverse
  - loops
authors: []
description: ''
externalLink: ''
series: []
editor_options: 
  chunk_output_type: console
---

## A Short General Primer on Loops

In my daily programming I barely use any base `for` loops. I used to rely on `lapply` 
and friends a lot, then switched to the `plyr` family of apply functions and eventually
ended up using the `purr::map_*` family of functions.

There is a bit of overhead involved in `map`and `[lvs]apply` can be faster / more efficient, 
but I do like the verbosity of `map` and friends as well as the possibility to use
lambda functions via `~`^[[R 4.1](https://cran.r-project.org/doc/manuals/r-release/NEWS.html) 
added also support for lambdas. Yet it requires some more typing and my current R Studio 
version `1.4.1103` does not recognize the syntax yet.]. 
Especially the predictability of the result as opposed to `sapply` for instance is a clear
plus. Also, if you want the apply function to return something else than a list the 
`map_*` variants become rather handy.

Compare:

```{r base-apply, error = TRUE}
## using the R 4.1. anon function syntax
(res <- lapply(1:3, \(x) x + 1))

## is typesafe but will always return a list, thus this will throw an error
sum(res)

## sapply can be a remedy
(res <- sapply(1:3, \(x) x + 1))
sum(res)

## however output format of sapply is not always the same
sapply(1:3, \(x) c(1, x)) ## a matrix
sapply(1:3, seq_len)      ## a list
sapply(1:3, \(x) x + 1)   ## a vector
```

to:

```{r purrr-apply, message = FALSE, error = TRUE}
library(purrr)

## map_* enforces you to return the right type

map_int(1:3, ~ .x + 1L)
map(1:3, seq_len)
map_dfr(1:3, ~ data.frame(x = .x))

## type safety is taken really seriously:
##   1 is a numeric, thus . + 1 is a numeric and __not__ an int
map_int(1:3, ~ .x + 1)
```

To conclude: `map` adds verbosity and type safety and allows for `tidyverse` anonymous 
functions. Enough reasons for me to completely rely on `map_` and friends^[`walk` for 
instance, when I am only interested in side effects.].

## Revisit [Sunburst Graphs](/2022/03/04/how-how-to-change-the-colors-in-a-sunburst-graph/)

In my post on how to color sunbursts graphs, I used a rather tedious and lengthy way of
bringing the data into the right format. We can use `purrr::accumulate` to generalize the 
idea while saving a lot of copy-pasted code:

```{r reduce-for-the-win, message = FALSE}
library(tidyr)
library(forcats)
library(dplyr)
titanic <- as_tibble(Titanic) %>% 
   mutate(across(where(is.character), fct_inorder))

prepare_sunburst_data <- function(data, ..., n = n(), total_name = "Total") {
   n <- enquo(n)
   grps <- enquos(...)
   data <- data %>% group_by(!!!grps) # [1]
   calc_margins <- function(.data) { # [3]
      grps <- .data %>% 
         groups()
      agg <- .data %>% 
         summarise(margin = sum(!!n), .groups = "drop") 
      if (length(grps)) {
         agg <- agg %>% 
            unite("id", !!!grps, remove = FALSE) %>% 
            mutate(label = !!grps[[1L]] %>% as.character())
         par_grps <- tail(grps, -1L)
         if (length(par_grps)) {
            agg <- agg %>% 
               unite("parent", !!!par_grps)
         } else {
            agg <- agg %>% 
               mutate(parent = total_name)
         }
      } else {
         agg <- agg %>% 
            mutate(label = total_name, id = total_name, parent = "")
      }
      agg %>% 
         select(id, parent, label, margin)
   }
   accumulate(grps, ~ .x %>% ungroup(!!.y), .init = data) %>%  # [2]
      map_dfr(calc_margins)
}

(sunburst_data <- prepare_sunburst_data(titanic, Survived, Sex, Age, Class, n = n))
```

How does the function work?

1. We use `enquos` to enquote all passed (via `...`) grouping variables and add these 
grouping variables to the data (`# [1]`).
1. Then, we use `accumulate` to create a list of data frames where each element uses one
grouping level *less*. That is int eh above example, the first element groups according to
`Survived`, `Sex`, `Age` and `Class`, the second one drops `Survived` the third one drops 
`Sex` and so on (`# [2]`).
1. Now, we can simply apply our `calc_margins` (`# [3]`) function, which uses the given 
summary function and creates meaningful ids and labels and sets the parent reference 
accordingly.

And as we are better in juding visuals rather than tables, here is the sunburst graph 
again to show that the data created by `prepare_sunburst_data` is indeed the same as the
approach by hand:

```{r sunburst}
library(plotly)
sunburst_data %>% 
   plot_ly() %>% 
   add_trace(ids = ~ id,
             labels = ~ label,
             parents = ~ parent,
             values = ~ margin,
             type = "sunburst",
             branchvalues = "total")
```


