---
title: How to Use Fully Qualified Tablenames in Dbplyr
author: Thorn Thaler
date: '2022-03-11'
slug: how-to-use-fully-qualified-tablenames-in-dbplyr
categories:
  - coding
tags:
  - dbplyr
  - database
  - SQL
authors: []
description: ''
externalLink: ''
series: []
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(dev.args = list(bg = "transparent"), 
                      fig.align = "center", message = FALSE)
```

## SQL Databases

I have only a working knowledge of databases and also more from a theoretical point of 
view rather than a deep understanding of the details of the respective implementation. 
That is, I know what normalization is (and why it is a good thing) and how to do `SELECT` 
or `JOIN` queries, but when it comes to the technical details of the different SQL 
flavors my knowledge dwindles away.

When I started my new job, I suddenly had to work with a huge ecosystem of databases and
I had to quickly catch up. We have a plethora of different databases hosted on different
servers and I learned that there is something like a `schema` and that tables (I rather 
should say objects, because it is not only tables) can (should?) be addressed by `[InstanceName].DatabaseName.SchemaName.ObjectName`. I also 
learned that sometimes SQL does not allow to specify the `DatabaseName` of this fully 
qualified address (try `CREATE db.schema.name` and you will get an error) and a thing or
two more about databases. 

As we are lots of colleagues working on the (Same) databases I started to use custom
schemata for my tables, procedures and views. I thought of schemata as something like a
namespace and depending on the SQL client you are using they are also nicely visually
grouped. Additionally, at some point our Database Administrators (DBAs) decided that the 
standard schema is **not** `dbo` anymore, but a personalized one. That is, if I do not 
specify the schema explicitly my objects are stored in the 
schema `OURDOMAIN\myusername`. 

I guess the reasoning for this change was to identify more easily who is the owner of an
object (or at least to make that visible to non DBAs), but of course it is a nightmare to
type and you have to quote the schema via `db.[OURDOMAIN\myusername].object`.

To make a long story short, I had to deal with schemata and databases from day 1 on. In
the beginning that was not a big deal, as I was primarily working with an SQL Client 
(SQL Server Management Studio (SSMS), admittedly a bit more than just a client). However, 
at some point I found the workflow a bit cumbersome to say the least:

1. Get data from SQL via SSMS.
1. Store it in a flat file.
1. Analyze the data in R.

Not only is it a pain in the neck (for me) to manipulate the data in SQL with the same 
efficiency as I would do in R, but also it is always incredibly difficult to gte data
out of SQL and into a flat file. Not even mentioning the problems which may arise from
stale data. Enters `DBI` abd `dbplyr`.

## Using R to Access Data

With [`DBI`](https://dbi.r-dbi.org/) as a general framework and
[`dbplyr`](https://dbplyr.tidyverse.org/) as a database backend for
[`dplyr`](https://dplyr.tidyverse.org/) I could marry the two worlds and I did not need
to leave my beloved RStudio environment for doing the end-to-end analysis.

I had, however, a rather steep learning curve (thanks 
[StackOverflow](https://stackoverflow.com/) 💖) until I figured how to address tables 
properly with their fully qualified name and I want to share some insights here:

```{r sql-dbplyr, eval = FALSE}
library(DBI)
library(dbplyr)
library(dplyr)

con <- dbConnect(odbc::odbc(), "my.database.server")

## 1. Use DBI::d 
tbl(con, Id(catalog = "mydatabase", schema = "myschema", table = "mytable"))

## actually names are ignored
tbl(con, Id(database = "mydatabase", middle_name = "myschema", object = "mytable"))

## as long as all are named
tbl(con, Id("mydatabase", "myschema", "mytable"))
## Error: All arguments to Id() must be named.

## We can check that it is all properly translated to
## <SQL>
## SELECT *
## FROM "mydatabase"."myschema"."mytable"

## 2. Use dbplyr::in_schema - however, need to provide database name 
## as part of "schema" and need to quote it
tbl(con, in_schema(sql("mydatabase.myschema"), "mytable"))

## this won't work
tbl(con, in_schema("mydatabase.myschema", "mytable"))

## the error message also shows why, if you do not escape it is seen as one schema
## <SQL> 
## SELECT *
## FROM "mydatabase.myschema"."mytable"


## 3. Use dbplyr::ident_q
tbl(con, ident_q("mydatabase.myschema.mytable"))

## note the _q version, ident would __not__ do
tbl(con, ident("mydatabase.myschema.mytable"))

## again compare the generated SQLS
## ident_q:
##   <SQL>
##   SELECT *
##   FROM mydatabase.myschema.mytable
## ident:
##   <SQL>
##   SELECT *
##   FROM "mydatabase.myschema.mytable"
```

To spice things a bit up, if you are using other querying functions like for instance
`DBI::dbReadTable` you do not have the same options, basically with `DBI` you have only
the  `Id` option:

```{r sql-dbi, eval = FALSE}
## works
dbReadTable(con, Id(catalog = "mydatabase", schema = "myschema", table = "mytable"))

## won't work
dbReadTable(con, ident_q("mydatabase.myschema.mytable"))

## neither does this
dbReadTable(con, in_schema(sql("mydatabase.myschema"), "mytable"))
```

Also be careful to have a recent `dbplyr` version (i.e. >= `2.1.1`) , where the `con` 
parameter is passed down to `as.sql` otherwise commands like 
`compute(..., in_schema(...), temporary = TRUE)`
[won't work](https://github.com/tidyverse/dbplyr/issues/595).

## Closing Remarks

Error messages are not always that straight forward to understand. However, whenever the
non working SQL is part of the error message, you should carefully look at where the
escaping puts `"` (for sQL Server, other backends quote differently). This gives you a 
hint, where things may have gone awry. Of course we can always fall back to pure SQL in
R (that is without relying on the abstraction layer of `dbplyr`), but that would be of 
courese only half the fun.